They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-20 paper/paper.tex
more methodology writing

The formula for the expected proportion of seriesn a best-of-\(s\)
format (\(s = 11\) rounds for CoD SnD), ending in \(r\) rounds
(\(r \leq s\)), \(\hat{\Phi}(r)\), for a constant round win probability
\(\phi\) (\(0 \leq \phi \leq 1\)) for one team\footnote{If
  \(\phi > 0.5\), then one might say that this team is the better team
  (known in hindsight).} is

\[
m = \frac{s + 1}{2}
\]

@@

\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(r - s)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m).
}\label{eq:series-length}\end{equation}

To evaluate the constant round win probability hypothesis--that is, that
the expected and observed round win rates, \(\hat{\Phi}(r)\) and
\(\Phi(r)\) respectively, are equal to one another--we can compute the
chi-square goodness of fit statistic, \(\chi^2\).

\begin{equation}\protect\hypertarget{eq:chi-squ}{}{
\chi^2 = \sum^R \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)}
}\label{eq:chi-squ}\end{equation}

For CoD SnD, \(r \in R = [6, 7, 8, 9, 10, 11]\).

@@

Removing our knowledge of a streak, we might model the Bernoulli round
win probability as

\begin{equation}\protect\hypertarget{eq:pwr}{}{
p^{w|r} = p(\text{win | r})=\{
\begin{array}{cl}
\frac{m}{r}, & \text{team wins series} \\
\frac{r - m}{r}, & \text{team loses series}
\end{array}.
}
\label{eq:pwr}
\end{equation}

The Bernoulli round loss probability after a streak of \(k\) round wins,
\(p^{\ell|kr}_0\), can be formulated symmetrically.

As shown by MS, we should expect the proportion of rounds wins
immediately following a streak of \(k\) rounds wins for a series lasting
\(r\) rounds, \(\hat{P}^{w|kr}_{MS}\), to be strictly less than
\(p^{w|r}\).\footnote{See Appendix E.1 of MS for details on the proof
  that follows.}

\begin{quote}
Theorem 1: \emph{Let}
\(\mathbf{X} = \{X_i\}^{n}_{i=1}, n \geq 3\)\emph{, be a sequence of
independent Bernoulli trials, each with probability of success}
\(0 < \pi < 1\)\emph{. Let} \(\hat{\Pi}_k(\mathbf{X})\) \emph{be the
proportion of successes on the subset of trials} \(I_k(\mathbf{X})\)
\emph{that immediately follow} \(k\) consecutive successes, that is,
\(\hat{\Pi}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{\Pi}_k\)
\emph{is a biased estimate of}
\(\mathbb{\Pi}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv p\)
\emph{for all} \(k\) \emph{such that} \(1 \leq k \leq n - 2\)\emph{. In
particular,}
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < \pi.\)
\end{quote}

MS note that there does not exists a closed form representation for
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]\) for
\(k > 1\). (\(\hat{\Pi}_k(\mathbf{X})\) is the analogue for
\(\hat{P}^{w|kr}_{MS}\) in my notation for CoD SnD.) Figure 1 shows how
the expected value varies as a function of the total number of trials
\(n\), given \(\pi\) and \(k > 1\).\footnote{I've adapted the code from
  Vafa (2017), which implements MS's framework (2018).}

\begin{figure}
\centering
\includegraphics{images/pwkr_ms.png}
\caption{The expected value of the proportion of successes on trials that immediately follow $k$ consecutive successes, $E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$, for Bernoulli trial success $\pi = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $n.$}
\end{figure}

My context is fundamentally different from that of GVT and MS, both of
whom focus on longitudinal data in controlled settings. GVT also perform
statistical tests on shots from players in live games,
i.e.~``observational'' data, but they note that their findings are
likely affected player shot selection in the face of defensive strategy
by the opposing team. The number of trials is fixed in their
experimental designs, but in CoD SnD, the number of rounds played is
determined as a function of the max number of possible rounds (\(s\))
and whether or not the team wins the series, as shown in Equation
\ref{eq:pwr}. The Bernoulli trial success probability, i.e.~a round win
in CoD SnD, is not independent of the opponent.

Consequently, a statistical test of the difference in \(\hat{P}^{w|kr}\)
and \(\hat{P}^{\ell|kr}\), as performed by GVT and MS to evaluate their
respective hypothesis regarding
\(E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]\) (Theorem
1), is not appropriate.

Let us consider another form of the expected proportion of rounds won
immediately after a streak of \(k\) round wins in a best-of-\(s\) series
given the length of the series (\(r\) rounds), the ``hindsight''
proportion \(\hat{P}^{w|kr}_0\). The Bernoulli round win probability
underlying \(\hat{P}^{w|kr}_0\) is

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0^{w|kr} = p_0(\text{win | k, r})=\{
\begin{array}{cl}
\frac{m - k}{i - k}, & \text{team wins series} \\
\frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}.
}
\label{eq:pwkr}
\end{equation}

A binomial test for the expected proportion of round wins following
streaks of \(k\) round wins in a series lasting \(r\) rounds,
\(\hat{P}^{w|kr}_0\), assuming \(p^{w|kr}_0\) as the probability of
success, can be used to verify that \(\hat{P}^{w|kr}_0\) is reasonable
as to use as a baseline with which to compare \(\hat{P}^{w|kr}_{MS}\).
If the binomial test shows that we cannot reject the null hypothesis
that the expected \(\hat{P}^{w|kr}_0\) is different from the observed
proportion \(P^{w|kr}\).

A proportion test of \(\hat{P}^{w|kr}_0\) and \(\hat{P}^{w|kr}_{MS}\)
can tell us whether the momentum exists in CoD SnD. A null result means
we cannot say that there is a difference between the hindsight expected
proportion \(\hat{P}^{w|kr}_0\) and the expected proportion adjusted for
streak-selection bias, \(\hat{P}^{w|kr}_0\).

>>>>

The formula for the expected proportion of series a best-of-\(s\) format
ending in \(r\) rounds (\(r \leq s\)), \(\hat{\Phi}(r)\), for a constant
round win probability \(\phi\) (\(0 \leq \phi \leq 1\)) for one
team\footnote{If \(\phi > 0.5\), then one might say that this team is
  the better team (known in hindsight).} is

@@

\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(r - s)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}

\(s = 11\) rounds for CoD SnD.\footnote{\(\phi\) and other symbols are
  selected so as to avoid reserve symbols, like \(p\) for usage in other
  contexts, without risking confusion. Upper-case symbol, e.g.~\(\Phi\),
  are consistently used in this paper to represent proportions, while
  lower-case symbols corresponding to proportions represent
  probabilities. ``Hats'', e.g.~\(\hat{\Phi}\), are used to convey
  expectations, while ``bare'' symbols mirroring hat-ed symbols convey
  observational data.}

To evaluate the constant round win probability null hypothesis---that
is, that the expected and observed round win rates, \(\hat{\Phi}(r)\)
and \(\Phi(r)\) respectively, are equal to one another---we can compute
the chi-square goodness of fit statistic, \(\chi^2\),

\begin{equation}\protect\hypertarget{eq:chi-squ}{}{
\chi^2 = \sum^R \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)}.
}\label{eq:chi-squ}\end{equation}

\(r \in R = [6, 7, 8, 9, 10, 11]\) for CoD SnD.

@@

Let us now consider round win rate immediately following a streak of
\(k\) wins, given that the series lasts \(r\) rounds. Removing our
knowledge of a streak, we might model the Bernoulli round win
probability as

\begin{equation}\protect\hypertarget{eq:pwr}{}{
p_0(\text{win} | r) = p^{w|r}_0 = \{
\begin{array}{rcll}
p^{w|r_W} & = & \frac{m}{r}, & \text{team wins series} \\
p^{w|r_L} & = & \frac{r - m}{r}, & \text{team loses series}
\end{array}
}
\label{eq:pwr}
\end{equation}

using \(m\) from Equation \ref{eq:m}. The Bernoulli round loss
probability for a series lasting \(r\) rounds, \(p^{\ell|kr}_0\)
(i.e.~\(\ell = 1 - w, w \in [0, 1]\)), can be formulated symmetrically.

As shown by MS, we should expect the proportion of rounds wins
immediately following a streak of \(k\) rounds wins for a series lasting
\(r\) rounds, \(\hat{P}^{w|kr}_{MS}\), to be strictly less than
\(p^{w|r}\).\footnote{See Appendix E.1 of MS for details on the proof
  that follows.}\footnote{Here, I choose symbols, e.g.~\(\pi\), \(x\),
  etc. that mostly do not conflict with syntax in other places. The
  exceptions are \(k\) and \(r\), which are used to represent streak
  length and number of trials globally.}

\begin{quote}
Theorem 1: \emph{Let}
\(\mathbf{X} = \{X_i\}^{r}_{i=1}, r \geq 3\)\emph{, be a sequence of
independent Bernoulli trials, each with probability of success}
\(0 < \pi < 1\)\emph{. Let} \(\hat{\Pi}_k(\mathbf{X})\) \emph{be the
proportion of successes on the subset of trials} \(I_k(\mathbf{X})\)
\emph{that immediately follow} \(k\) consecutive successes, that is,
\(\hat{\Pi}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{\Pi}_k\)
\emph{is a biased estimate of}
\(\mathbb{P}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv \pi\)
\emph{for all} \(k\) \emph{such that} \(1 \leq k \leq r - 2\)\emph{. In
particular,}
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < \pi.\)
\end{quote}

MS notes that there does not exists a closed form representation for
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]\),
henceforth \(E^\Pi\), for \(k \> 1\). One may recursively build a
collection of dictionaries to quantify \(E^\Pi\) analytically, or, more
reasonably for large values of \(n\), one may run simulations to
estimate the expected value. Indeed, Figure 1 uses simulation to
estimate \(E^\Pi\) as a function of the total number of trials \(r\),
given \(\pi\) and \(k\).\footnote{I've adapted the code from Vafa
  (2017), which implements MS's framework (2018).}\footnote{Note that
  \(\pi\) and \(\hat{\Pi}_k(\mathbf{X})\) are analogues for
  \(p_0^{w|r}\) and \(\hat{P}^{w|kr}_{MS}\) in my notation for CoD SnD.}

\begin{figure}
\centering
\includegraphics{images/pwkr_ms.png}
\caption{The expected value of the proportion of successes on trials, $E^\Pi$, that immediately follow $k$ consecutive successes for Bernoulli trial success $\pi = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $r$.}
\end{figure}

My context is fundamentally different from that of GVT and MS, both of
whom focus on longitudinal data in controlled settings. GVT also
performs statistical tests on shots from players in live games,
i.e.~``observational'' data, but they note that their findings are
likely affected player shot selection in the face of defensive strategy
by the opposing team. The number of trials is fixed in their
experimental designs, but in CoD SnD, the number of rounds played is
determined as a function of the max number of possible rounds (\(s\))
and whether or not the team wins the series, as shown in Equation
\ref{eq:pwr}. The Bernoulli trial success probability, i.e.~a round win
in CoD SnD, is not independent of the opponent. Consequently, a
statistical test of the difference in \(\hat{P}^{w|kr}\) and
\(\hat{P}^{\ell|kr}\), as performed by GVT and MS to evaluate their
respective hypothesis regarding post-streak success rates is not
appropriate.

One might consider another form of the expected proportion of rounds won
immediately after a streak of \(k\) round wins in a best-of-\(s\) series
given the length of the series (\(r\) rounds), the ``notional''
proportion \(\hat{P}^{w|kr}_0\). The Bernoulli round win probability
underlying \(\hat{P}^{w|kr}_0\) is

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{w|kr}_0 =\{
\begin{array}{rcll}
p_0^{w|kr_W} & = & \frac{m - k}{i - k}, & \text{team wins series} \\
p_0^{w|kr_L} & = & \frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}

using \(m\) from Equation \ref{eq:m}. We can perform a binomial test to
test the null hypothesis \(H_0 : \omega = \omega_0\) for the observed
probability of success \(\omega\) and a user-specified \(\omega_0\),
where \(0 \leq \omega_0 \leq 1\).\footnote{\(\omega\) and other symbols
  are chosen so as to not conflict with symbols already used elsewhere
  for in other equations, either generic or for CoD SnD specifically.}

If there are \(\sigma\) observed successes in a sample of \(r\) trials
and we expect that there should be \(r * \omega_0\), the probability of
getting this expected number of successes is

\begin{equation}\protect\hypertarget{eq:binom}{}{
\Pr(\sigma) = {\binom {r}{\sigma}} \omega^{\sigma}(1-\omega)^{r-\sigma}.
}\label{eq:binom}\end{equation}

Plugging in the notional probability \(p^{w|kr}_0\) for the null
\(\omega_0\) and the observed probability
\(p^{w|kr} = \frac{N^{w|kr}}{P^{w|kr}}\), where \(N^{w|kr}\) is the
observed number of instances where a teams wins the round immediately
following a \(k\)-round winning streak in a series lasting \(r\) rounds,
for the observed \(\omega\), we can evaluate whether the observed
probability is not significantly different from the notional
probability. If we reject this hypothesis, then we might consider team
momentum, represented by post-streak success, plausible.

Returning to the streak-selection adjustment framework proposed by MS,
one can perform the same binomial test, swapping out \(p^{w|kr}_0\) with
\(p^{w|kr}_{MS}\). Of course, given the caveats mentioned before, the
results should be heeded with caution.
